SHARP II: Dark Matter Substructure, Gravitational Lens Flux Ratio Anomalies, and the Case of B1555+375: Beware the Effect of Edge-on Disks

### 2015-04-27 papers/B1555_EdgeOn/B1555_letter.tex
change to text

A common feature of simulations of structure formation is the presence
of thousands of subhalos that are associated with larger mass aloes \citep[e.g.][]{Springel08}.
However, observations of the Local Group find many fewer satellite
galaxies than are predicted by the simulations, even when taking into
account completeness corrections arising from the limited sky coverage
imposed by the Galactic Plane and the latest discovery of new dwarf galaxies \citep{DES15,Kop15}.  
This is the famous ``missing satellite'' problem \citep{Klypin1999, Moore1999, S07}. In order to
understand this discrepancy, it is imperative to build up a large
sample of satellites/subhalos in galaxies outside the Local Group.
 However, this process is challenging due to the faintness of the satellite
galaxies.  Furthermore, lower-mass satellites may not be able to
retain the gas needed for ongoing star formation \citep[e.g.,][]{P11},
rendering them effectively dark.  These factors make gravitational
lensing a powerful and promising tool for the detection of
substructure.

It was first suggested by \citet{Mao1998} that flux ratio anomalies
observed in radio loud multiple images of lensed quasar could be
interpreted as a telltale sign of the presence of substructure in lens
galaxies.  Indeed, small perturbations in the gravitational potential
are sufficient to cause flux ratio anomalies in strong lensed systems
\citep{Bradac02}.  Moreover, unlike in the optical or infrared, in the
radio band, lensed images are not sensitive to microlensing from
stars nor dust extinction. This makes flux ratio anomalies a promising
tool to detect \textbf{and measure} dark matter substructures \citep{Dalal2002, N13}.  At
present, the amount of substructure derived from flux ratio anomalies
is in disagreement with predictions from numerical simulations, with
the former requiring a larger fraction of mass in substructure than
expected \citep{Xu14}. While this could be the result of small sample
statistics, it could also be an indication that flux ratio anomalies
have a different origin than substructure.
%
An alternative method for
the detection of substructure in gravitational lens galaxies based on
the surface brightness distribution of extended arcs and Einstein rings
has been introduced by \citet{K05} and \citet{V09}. Interestingly, the fraction of
substructure measured with this \emph{gravitational imaging technique}
is currently in agreement with theoretical expectations \citep{V14a}.

@@

As a first trial of modeling, we create a single SIE model to check
the performance of a simple lensing potential. Consistent with
previous studies \citep{Marlow99, Xu14}, our single elliptical
potential model cannot predict the flux anomaly in B1555+375.  We then
create two separate models of the lensing potential that both
incorporate the disk.  In the first we treat the lens as a SIE bulge
and an exponential edge-on disk (SIE+expdisk).  In the second model
the disk is represented as a very elongated SIE component (2-SIE).
The lens system is red, with a strong jump in flux between the F814W
(roughly I) and F160W (roughly H) bands (Fig.~\ref{fig:multiband}), so
we assign a lens redshift of $z_\ell = 1.0$ and a source redshift of
$z_s =1.5$. Our steps to explore the parameter space are: (1) Apply
strong priors from SHARP AO image to the edge-on disk position angle
and ellipticity, and use the best-fit model of \citet{Marlow99} as the
starting point for the bulge component and the source position. (2)
Let all parameters without priors vary to reach a temporary best
fit. (3) Start with the best fit in step 2 and let all parameters with
priors vary to reach another best fit (also ensure this best fit is
not far from the priors). (4) Let all parameters vary to reach the
final optimal model.  The full best-fit parameters are shown in Table
2. Both of these models produce lensed images with positions and
fluxes that are good matches to the radio observations.  While both of
the two component models produce significantly improved matches to the
fluxes than the single SIE model, the 2-SIE version has a somewhat
better performance than the SIE+expdisk model in reproducing flux
ratios. The comparison of the observations and our lens modeling is
given in Figure 3 and Tables 3 \& 4.  We have not included an external
shear component into our model because the constraints provided by the
image positions and fluxes would not be sufficient to constrain such a
model. However, we note that the image not affected by dust (image C)
in the Keck AO data is consistent with negligible shear (i.e. shear
strength $\Gamma_s=0.002$), when modeled with the pixelized technique
of \citet{V09}.

\begin{figure}
%\plotone{1555_ao_merlin_overlay.eps}
\includegraphics[width=84mm]{1555_ao_merlin_overlay.eps}
\caption{High-resolution  $K^\prime$ band imaging of the B1555+375 lens system 
with contours from the MERLIN observation of \citet{Marlow99} overlaid.
%
\label{fig:merlin}}
\end{figure}

@@

\begin{table}
% \centering
% \begin{minipage}{140mm}
  \caption{Best-fit parameters of B1555+375 lens models.}
  \begin{tabular}{@{}ccc}
\hline 
 Parameter  & \multicolumn{2}{c}{Lens model} \\
		&SIE+expdisk& 2-SIE		   
\\
\hline
$x_1$  	  & $-0.1883$	& $-0.1813$  \\
$y_1$	  &$-0.1923$	&$-0.1695$  \\
$x_2$	  &$-0.1615$ 	&$-0.1593$  \\
$y_2$	  &$-0.2502$	& $-0.1843$  \\
$\sigma_1$ &$141.3$     & $144.9$ \\
$e_1$	  & $0.27$	& $0.27$ \\
$\theta_1$ &$105.5 \degr$ & $97.7 \degr$ \\
$M_{tot,2}$  & $3.59\times 10^{10} $  & ...	 \\  
$\sigma_2$ & ...        &$123.5$ \\  
$e_2$	  &$0.84$	&$0.83$  \\
$\theta_2$ &$7.4\degr$ &$7.4\degr$  \\
$r_e$	  & $0.20 ''$ &  ... \\
\hline
$\chi ^2$ & $1.05$ & $0.04$ \\
\hline
\end{tabular}

%\end{minipage}
\medskip
Subscripts 1 and 2 represent bulge and disk components in each model respectively. Positions are offsets from radio image component A measured in units of arcsec. The velocity dispersion $\sigma$ is in units of $km/s$ and disk mass $M_{tot}$ is in units of $h^{-1} M_{\odot}$. $e$ is ellipticity and $\theta$ is the position angle measured east of north.

@@

%\end{minipage}
\medskip
Comparison between radio data and model-predicted flux ratios. The radio observation data is quoted from Tables 2 and Table 3 in \citet{Marlow99}. 1-SIE model is quoted from Table 5 in \citet{Marlow99}.

@@

In this letter, we focus on the effect of more complicated mass
distributions for the lens galaxy on flux ratio anomalies.
Traditionally, flux anomalous lensed quasars are radio loud systems
that are firstly discovered in radio observations. While radio data
provide reliable information on the positions and the fluxes of the
lensed images, they are often unable to detect the lens
galaxies. Moreover, lens modeling based only on the positions of the
four images can be highly degenerate and prior assumptions play a key
role in the interpretation of the observed flux ratio anomaly.  Since
lens galaxies are typically at redshifts of $z \leq 1$, optical and
near-infrared observations are ideal for measuring the light distribution
of lens galaxies and obtaining useful prior information on the underlying
mass distributions. Indeed \citet{SHARP12} showed that high
resolution IR imaging can help to significantly improve the precision
of the lens modeling. This improvement is even more dramatic in the
case of B1555+375, where HST and SHARP AO high resolution imaging has
allowed us to detect, for the fist time, the presence of an edge-on
disk. This new discovered edge-on disk provides a strong prior for our
lens modeling.  With a bulge plus an edge-on disk, our models produce
matches to both the radio image positions and their flux ratio, while
models with only one SIE component \citep{Marlow99,Xu14} fail to
reproduce the fluxes correctly. Our successful modeling indicates
that the kiloparsec-scale structure of the lens galaxy itself is
sufficient to explain the flux ratio anomalies in B1555+375, without
any need to invoke the presence of substructure.

>>>>

A common feature of simulations of structure formation is the presence
of thousands of subhalos that are associated with larger mass haloes \citep[e.g.][]{Springel08}.
However, observations of the Local Group find many fewer satellite
galaxies than are predicted by the simulations, even when taking into
account completeness corrections arising from the limited sky coverage
imposed by the Galactic Plane and the latest discovery of new dwarf galaxies \citep{DES15,Kop15}.  
This is the famous ``missing satellite'' problem \citep{Klypin1999, Moore1999, S07}. In order to
understand this discrepancy, it is imperative to build up a large
sample of satellites/subhalos in galaxies outside the Local Group.
 However, this process is challenging due to the faintness of the satellite
galaxies.  Furthermore, lower-mass satellites may not be able to
retain the gas needed for ongoing star formation \citep[e.g.,][]{P11},
rendering them effectively dark.  These factors make gravitational
lensing a powerful and promising tool for the detection of
substructure.

It was first suggested by \citet{Mao1998} that flux ratio anomalies
observed in radio loud multiple images of lensed quasars could be
interpreted as a telltale sign of the presence of substructure in lens
galaxies.  Indeed, small perturbations in the gravitational potential
are sufficient to cause flux ratio anomalies in strong lensed systems
\citep{Bradac02}.  Moreover, unlike in the optical or infrared, in the
radio band, lensed images are not sensitive to microlensing from
stars nor dust extinction. This makes flux ratio anomalies a promising
tool to detect \textbf{and measure} dark matter substructures \citep{Dalal2002, N13}.  At
present, the amount of substructure derived from flux ratio anomalies
is in disagreement with predictions from numerical simulations, with
the former requiring a larger fraction of mass in substructure than
expected \citep{Xu14}. While this could be the result of small sample
statistics, it could also be an indication that flux ratio anomalies
have a different origin than substructure.
%
An alternative method for
the detection of substructure in gravitational lens galaxies based on
the surface brightness distribution of extended arcs and Einstein rings
has been introduced by \citet{K05} and \citet{V09}. Interestingly, the fraction of
substructure measured with this \emph{gravitational imaging technique}
is currently in agreement with theoretical expectations \citep{V14a}.

@@

As a first trial of modeling, we create a single SIE model to check
the performance of a simple lensing potential. Consistent with
previous studies \citep{Marlow99, Xu14}, our single elliptical
potential model cannot predict the flux anomaly in B1555+375.  We then
create two separate models of the lensing potential that both
incorporate the disk.  In the first we treat the lens as a SIE bulge\textbf{/halo}
and an exponential edge-on disk (SIE+expdisk).  In the second model
the disk is represented as a very elongated SIE component (2-SIE).
The lens system is red, with a strong jump in flux between the F814W
(roughly I) and F160W (roughly H) bands (Fig.~\ref{fig:multiband}), so
we assign a lens redshift of $z_\ell = 1.0$ and a source redshift of
$z_s =1.5$. \textbf{This assumption on redshifts is due to that some lens model parameters contain distance information. Our choice of redshifts does not affects the predicted magnifications nor flux ratios.} Our steps to explore the parameter space are: (1) Apply
strong priors from the SHARP AO image to the edge-on disk position angle
and ellipticity, and use the best-fit model of \citet{Marlow99} as the
starting point for the bulge\textbf{/halo} component and the source position. (2)
Let all parameters without priors vary to reach a temporary best
fit. (3) Start with the best fit in step 2 and let all parameters with
priors vary to reach another best fit (also ensure this best fit is
not far from the priors). (4) Let all parameters vary to reach the
final optimal model.  The full best-fit parameters are shown in Table
2. Both of these models produce lensed images with positions and
fluxes that are good matches to the radio observations.  While both of
the two component models produce significantly improved matches to the
fluxes than the single SIE model, the 2-SIE version has a somewhat
better performance than the SIE+expdisk model in reproducing flux
ratios. The comparison of the observations and our lens modeling is
given in Figure 3 and Tables 3 \& 4.  We have not included an external
shear component into our model because the constraints provided by the
image positions and fluxes would not be sufficient to constrain such a
model. However, we note that the image not affected by dust (image C)
in the Keck AO data is consistent with negligible shear (i.e. shear
strength $\Gamma_s=0.002$), when modeled with the pixelized technique
of \citet{V09}.

\begin{figure}
%\plotone{1555_ao_merlin_overlay.eps}
\includegraphics[width=84mm]{1555_ao_merlin_overlay.eps}
\caption{High-resolution  $K^\prime$ band imaging of the B1555+375 lens system 
with contours from the MERLIN observation of \citet{Marlow99} overlaid. \textbf{From left to right, four components are A, B (merging doubles), D(lowest spot), and C(right).}
%
\label{fig:merlin}}
\end{figure}

@@

\begin{table}
% \centering
% \begin{minipage}{140mm}
  \caption{Best-fit parameters of B1555+375 lens models.}
  \begin{tabular}{@{}ccc}
\hline 
 Parameter  & \multicolumn{2}{c}{Lens model} \\
		&SIE+expdisk& 2-SIE		   
\\
\hline
$x_1$  	  & $-0.1883$	& $-0.1813$  \\
$y_1$	  &$-0.1923$	&$-0.1695$  \\

$\sigma_1$ &$141.3$     & $144.9$ \\
$e_1$	  & $0.27$	& $0.27$ \\
$\theta_1$ &$105.5 \degr$ & $97.7 \degr$ \\
\hline
$x_2$	  &$-0.1615$ 	&$-0.1593$  \\
$y_2$	  &$-0.2502$	& $-0.1843$  \\
$M_{tot,2}$  & $3.59\times 10^{10} $  & ...	 \\  
$\sigma_2$ & ...        &$123.5$ \\  
$e_2$	  &$0.84$	&$0.83$  \\
$\theta_2$ &$7.4\degr$ &$7.4\degr$  \\
$r_e$	  & $0.20 ''$ &  ... \\
\hline
$\chi ^2$ & $1.05$ & $0.04$ \\
\hline
\end{tabular}

%\end{minipage}
\medskip
Subscripts 1 and 2 represent bulge\textbf{/halo} and disk components in each model respectively. Positions are offsets from radio image component A measured in units of arcsec. The velocity dispersion $\sigma$ is in units of $km/s$ and disk mass $M_{tot}$ is in units of $h^{-1} M_{\odot}$. $e$ is ellipticity and $\theta$ is the position angle measured east of north.

@@

%\end{minipage}
\medskip
Comparison between radio data and model-predicted flux ratios. The radio observation data is quoted from Tables 2 and Table 3 in \citet{Marlow99}. 1-SIE model is quoted from Table 5 in \citet{Marlow99}. [WE HAVE 1-SIE MODEL NOW. SHOULD I QUOTE FROM MARLOW OR OUR OWN?--JW]

@@

In this letter, we focus on the effect of more complicated mass
distributions for the lens galaxy on flux ratio anomalies.
Traditionally, flux anomalous lensed quasars are radio loud systems
that are firstly discovered in radio observations. While radio data
provide reliable information on the positions and the fluxes of the
lensed images, they are often unable to detect the lens
galaxies. Moreover, lens modeling based only on the positions of the
four images can be highly degenerate and prior assumptions play a key
role in the interpretation of the observed flux ratio anomaly.  Since
lens galaxies are typically at redshifts of $z \leq 1$, optical and
near-infrared observations are ideal for measuring the light distribution
of lens galaxies and obtaining useful prior information on the underlying
mass distributions. Indeed \citet{SHARP12} showed that high
resolution IR imaging can help to significantly improve the precision
of the lens modeling. This improvement is even more dramatic in the
case of B1555+375, where HST and SHARP AO high resolution imaging has
allowed us to detect, for the fist time, the presence of an edge-on
disk. This newly discovered edge-on disk provides a strong prior for our
lens modeling.  With \textbf{an additional disk component}, our models produce
matches to both the radio image positions and their flux ratio, while
models with only one SIE component \citep{Marlow99,Xu14} fail to
reproduce the fluxes correctly. Our successful modeling indicates
that the kiloparsec-scale structure of the lens galaxy itself is
sufficient to explain the flux ratio anomalies in B1555+375, without
any need to invoke the presence of substructure.
